---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: pdf_document
authors: Kristie Semanchik 
date: 14 Sep 2024
---
# Required Packages 
To begin this tutorial, please load these required packages.  
```{r Load Libraries, include=FALSE}
if (!require("datasets")) install.packages("datasets"); library(datasets)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
if (!require("dbplyr")) install.packages("dbplyr"); library(dbplyr)
if (!require("ggplot2")) install.packages("ggplot2"); library(ggplot2)
```

# Objectives

1. By the end of this tutorial, students will be able to understand and confidently apply basic data analysis techniques in RStudio, including statistical tests ranging from summary statistics to ANOVAs, to interpret and analyze data sets.

1.1 Descriptive Statistics: Students will be able to calculate basic descriptive statistics including summary statistics, such as sample size, min, max, mean, median, and standard deviation. 

1.2: Statistical Testing: Students will be able perform statistical tests including two-sample t-test, one way ANOVA, and a two way ANOVA, as well as understand the outputs to make informed decisions about the data.  

2. By the end of this tutorial, students should be able to confidently construct different graphs, including but not limited to, scatterplots, bar charts, box and whisker, histograms, QQ plots, etc. Students will have clear interpretations of these graphs and how they represent relationships in the data.

2.1 Data Visualization: Students will be able to generate visualizations of data using the ggplot function and choose the appropriate graph according to relationships between variables, types of variables in the dataset, and your hypothesis.

2.2 Graphical Analysis: Students will be able to effectively analyze these graphs and have the tools to understand what the visuals mean in terms of their hypothesis and/or question. Students will maintain clarity in explaining what these graphs do, what our objective is in making these graphs, and how this brings us closer to our final goal.

# Section 1.1: Descriptive Statistics
## Summary Statistics 
Summary statistics are used to provide a quick summary of the data! They are particularly useful to understand, communicate, and compare the data. 

We are going to learn how to calculate each statistic individually, and then we will learn how to group the statistics together.  

### Example
The statistics we are most interested in are min, max, mean, median, and standard deviation. 

We will begin by calculating these statistics for overall uptake. 
```{r}
data("CO2")
min(CO2$uptake,na.rm=TRUE)
max(CO2$uptake,na.rm=TRUE)
mean(CO2$uptake,na.rm=TRUE)
median(CO2$uptake,na.rm=TRUE)
sd(CO2$uptake,na.rm=TRUE)

```
There is an easier way to group these statistics together that allows us to look at the statistics for type and treatment individually! This requires us using a function called group_by.

We will now calculate the summary statistics grouped by type. 
```{r}
summary_stats_type<- CO2 |>
  group_by(Type) |>
  summarise(count_uptake=n(),
            min_uptake=min(uptake), 
            max_uptake=max(uptake),
            mean_uptake=mean(uptake),
            median_uptake=median(uptake), 
            sd_uptake=sd(uptake))
print(summary_stats_type)
```
The mean uptake for plants from Quebec is 33.54, whereas the mean uptake for plants from Mississippi is 20.88. These numbers seem different, but we cannot confirm significance until we run statistical tests (Section 1.2)! 

### Practice 
It is now your chance to calculate the summary statistics grouped by Treatment. 
```{r}
# perform summary statistics here! 
```
What is the mean uptake for chilled? What about for nonchilled? Do you think there is a difference? 

# Section 1.2: Statistical Testing

## Statistical Test: t-test
A t-test is used to test if the means of two independent variables are significantly different. 

### Example
We will begin by performing a two-sample t-test to compare CO2 uptake between the two types of plants (Quebec and Mississippi). 

The formula used is t.test(dependent variable ~ independent variable, data= data frame)
Dependent Variable: uptake
Independent Variable: type
```{r}
t.test(uptake~Type, data=CO2)
```
The output shows us the p-value, which will tell us whether the difference in means is statistically significant. If the p-value is less than 0.05, we conclude that there is a statistically significant difference in CO2 uptake between Quebec and Mississippi.

The p-value is 4.451e-09. We can conclude that there is a statistically significant difference in CO2 uptake between plants from Quebec and Mississippi. 

### Practice
It is now your chance to practice performing a two-sample t-test to compare CO2 uptake between the two treatment types (chilled and nonchilled). 

What are the dependent and independent variables?
Dependent Variable: 
Independent Variable: 
```{r}
# perform a two-sample t-test here 
```
What was the p-value?

What does this tell us about the relationship between CO2 uptake and treatment? 

## Statistical Test: one way ANOVA
A one-way analysis of variance (ANOVA) is used to determine if the means more than 2 independent treatment groups are significantly different. 

This data set only has 2 treatment groups per independent variable -- that's okay! The results of a one-way ANOVA will be equivalent to a two-sample t-test when analyzing 2 treatment groups.  

### Example
We will begin by performing a one-way ANOVA to compare CO2 uptake between the two types of plants (Quebec and Mississippi). 

The formula used is name<-aov(dependent variable ~ independent variable, data= data frame)
Dependent Variable: uptake
Independent Variable: type
```{r}
type_anova<-aov(uptake~Type, data=CO2)
summary(type_anova)
```
The output shows us the Pr(>F), or the p-value, which will tell us whether the difference in means is statistically significant. Again, if the p-value is less than 0.05, we conclude that there is a statistically significant difference in CO2 uptake between Quebec and Mississippi.

The p-value is 3.83e-09. We can conclude that there is a statistically significant difference in CO2 uptake between plants from Quebec and Mississippi.

### Practice 
It is now your chance to practice performing a one way ANOVA to compare CO2 uptake between the two treatment types (chilled and nonchilled). 

What are the dependent and independent variables?
Dependent Variable: 
Independent Variable: 
```{r}
# perform one-way ANOVA here!
```
What was the p-value?

What does this tell us about the relationship between CO2 uptake and treatment? 

## Statistical Test: two way ANOVA
A two-way analysis of variance (ANOVA) is used to determine how two independent variables, in combination, affect a dependent variable. 

While this is a more advanced statistical test, there is no example, just practice! But don't worry, we will walk you through the code! 

### Practice 
The formula used is:
name<-aov(dependent variable ~ independent variable1 * independent variable2, data= data frame)

Based on what we have been doing, what are our dependent and independent variables?
Dependent: 
Independent 1:
Independent 2:
*note that independent variables 1 and 2 can be interchangeable! 
```{r}
try<-aov(uptake ~ Treatment * Type, data= CO2)
summary(try)
```
What was the p-value for treatment? 

What was the p-value for type? 

What was the p-value for the interaction between treatment and type? 

What is a two-way ANOVA used for? What do these results tell us about our two independent variables? 

#Section 2.1 Data Visualization
Data visualization is the graphical representation of information and/or data. The integration of visual elements into you data analysis is crucial for not only your understanding, but also your audience's understanding of patterns in the data. Data visualization also provides easier interpretation and highlights outliers or abnormalities in the data. 

Prior to beginning our data visualization, we must form a hypothesis. Do we think that the treatment type (chilled or non chilled conditions) will have a significant impact on CO2 uptake? Do we think that the type of plant (Quebec or Mississippi variant) will have a significant impact on CO2 uptake? Here is a hypothesis we can use for this tutorial

We hypothesize that the plant variant from Quebec will have a statistically significant difference in CO2 uptake from the Mississippi variant and that there will be higher CO2 uptake in the Quebec variant. 

Now you try! State a hypothesis about the levels of CO2 uptake dependent on the treatment of plants. 

##Histogram
We can use a histogram to understand and determine the distribution of the data. Let's start by analyzing the CO2 uptake depending on the type of plants (Quebec and Mississippi variants)


```{r}
ggplot(CO2) +
  aes(x = uptake, fill = Type) +
  geom_histogram(bins=100) +
  theme_cowplot()
```
##Practice Histogram
Now you can try to use this code to create a histogram comparing CO2 uptake in different treatment types (chilled and non-chilled treatments)

```{r}

```

##Boxplot
Because we are comparing numerical and categorical variables, it would be favorable to use a boxplot or a bar chart to visualize the data. If we were comparing numerical varibles to other numerical variables, it would be more favorable to use a scatterplot or line plot. Let's start off by plotting the uptake based on the type of plant using a scatterplot. 


```{r}
ggplot(CO2, aes(x = uptake, y = Type, color = Type)) +
    geom_boxplot() + geom_jitter() +theme_cowplot()

```
Based on this graph, it appears that the Quebec variant has a higher mean, median, Q1 and Q3. However, we cannot deduce this just by looking at the graph, we must complete more statistical testing.

###Practice Scatterplot
Now you can try to create a scatterplot of CO2 uptake based on treatment administered!
```{r}


```
For the sake of the tutorial, lets create a bar chart of the same variables in the boxplot, just to compare the two. 
##Bar Chart
```{r}
ggplot(CO2, aes(x = Type, y=uptake), fill = Type) + geom_bar(stat = "summary", position = "dodge", color = "black")+ theme_cowplot() + scale_fill_grey()+ xlab("Type of Plant") + ylab("CO2 Uptake")

```
The bar chart yields the (seemingly) same results as the boxplot. This time, we used the xlab and ylab codes to add specific axis labels. 

###Practice Bar Chart 

Try to make a bar plot of uptake vs treatment!
```{r}

```

#Section 2.2: Graphical Analysis
##Linear Models
We can now analyze the relationship between uptake and type of plant using linear models. We must first convert out categorical variable (type) to a factor so that it will be recognized in the linear model


```{r}
CO2$Type <- as.factor(CO2$Type)

lmType <- lm(uptake ~ Type, data = CO2)
summary(lmType)
```
This linear model displays that there is a statistically significant difference between the uptakes of the Quebec and Mississippi variants because we observe a p-value of 3.83 x 10^-9 (lower than .05). We can see at the estimate tab that TypeMississippi has a value of -12.660. This means that for every unit of CO2 uptake observed in the Quebec variant, the Mississippi variant is 12.660 units lower. Therefore, our hypothesis regarding the value of the Quebec variant versus the Mississippi variant is supported. 
###Practice Linear Models
Now try this with the uptake vs treatment! Remember to convert treatment to a factor!!
```{r}

```


What was your p-value?

What does your p- value say about the statistical significance of our findings?

Was your hypothesis proven or disproven?









