---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: pdf_document
authors: Kristie Semanchik 
date: 14 Sep 2024
---
# Required Packages 
To begin this tutorial, please load these required packages.  
```{r Load Libraries, include=FALSE}
if (!require("datasets")) install.packages("datasets"); library(datasets)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
if (!require("dbplyr")) install.packages("dbplyr"); library(dbplyr)
if (!require("ggplot2")) install.packages("ggplot2"); library(ggplot2)
```

# Objectives

1. By the end of this tutorial, students will be able to understand and confidently apply basic data analysis techniques in RStudio, including statistical tests ranging from summary statistics to ANOVAs, to interpret and analyze data sets.

1.1 Descriptive Statistics: Students will be able to calculate basic descriptive statistics including summary statistics, such as sample size, min, max, mean, median, and standard deviation. 

1.2: Statistical Testing: Students will be able perform statistical tests including two-sample t-test, one way ANOVA, and a two way ANOVA, as well as understand the outputs to make informed decisions about the data.  


# Section 1.1: Descriptive Statistics
## Summary Statistics 
Summary statistics are used to provide a quick summary of the data! They are particularly useful to understand, communicate, and compare the data. 

We are going to learn how to calculate each statistic individually, and then we will learn how to group the statistics together.  

### Example
The statistics we are most interested in are min, max, mean, median, and standard deviation. 

We will begin by calculating these statistics for overall uptake. 
```{r}
data("CO2")
min(CO2$uptake,na.rm=TRUE)
max(CO2$uptake,na.rm=TRUE)
mean(CO2$uptake,na.rm=TRUE)
median(CO2$uptake,na.rm=TRUE)
sd(CO2$uptake,na.rm=TRUE)

```
There is an easier way to group these statistics together that allows us to look at the statistics for type and treatment individually! This requires us using a function called group_by.

We will now calculate the summary statistics grouped by type. 
```{r}
summary_stats_type<- CO2 |>
  group_by(Type) |>
  summarise(count_uptake=n(),
            min_uptake=min(uptake), 
            max_uptake=max(uptake),
            mean_uptake=mean(uptake),
            median_uptake=median(uptake), 
            sd_uptake=sd(uptake))
print(summary_stats_type)
```
The mean uptake for plants from Quebec is 33.54, whereas the mean uptake for plants from Mississippi is 20.88. These numbers seem different, but we cannot confirm significance until we run statistical tests (Section 1.2)! 

### Practice 
It is now your chance to calculate the summary statistics grouped by Treatment. 
```{r}
# perform summary statistics here! 
```
What is the mean uptake for chilled? What about for nonchilled? Do you think there is a difference? 

# Section 1.2: Statistical Testing

## Statistical Test: t-test
A t-test is used to test if the means of two independent variables are significantly different. 

### Example
We will begin by performing a two-sample t-test to compare CO2 uptake between the two types of plants (Quebec and Mississippi). 

The formula used is t.test(dependent variable ~ independent variable, data= data frame)
Dependent Variable: uptake
Independent Variable: type
```{r}
t.test(uptake~Type, data=CO2)
```
The output shows us the p-value, which will tell us whether the difference in means is statistically significant. If the p-value is less than 0.05, we conclude that there is a statistically significant difference in CO2 uptake between Quebec and Mississippi.

The p-value is 4.451e-09. We can conclude that there is a statistically significant difference in CO2 uptake between plants from Quebec and Mississippi. 

### Practice
It is now your chance to practice performing a two-sample t-test to compare CO2 uptake between the two treatment types (chilled and nonchilled). 

What are the dependent and independent variables?
Dependent Variable: 
Independent Variable: 
```{r}
# perform a two-sample t-test here 
```
What was the p-value?

What does this tell us about the relationship between CO2 uptake and treatment? 

## Statistical Test: one way ANOVA
A one-way analysis of variance (ANOVA) is used to determine if the means more than 2 independent treatment groups are significantly different. 

This data set only has 2 treatment groups per independent variable -- that's okay! The results of a one-way ANOVA will be equivalent to a two-sample t-test when analyzing 2 treatment groups.  

### Example
We will begin by performing a one-way ANOVA to compare CO2 uptake between the two types of plants (Quebec and Mississippi). 

The formula used is name<-aov(dependent variable ~ independent variable, data= data frame)
Dependent Variable: uptake
Independent Variable: type
```{r}
type_anova<-aov(uptake~Type, data=CO2)
summary(type_anova)
```
The output shows us the Pr(>F), or the p-value, which will tell us whether the difference in means is statistically significant. Again, if the p-value is less than 0.05, we conclude that there is a statistically significant difference in CO2 uptake between Quebec and Mississippi.

The p-value is 3.83e-09. We can conclude that there is a statistically significant difference in CO2 uptake between plants from Quebec and Mississippi.

### Practice 
It is now your chance to practice performing a one way ANOVA to compare CO2 uptake between the two treatment types (chilled and nonchilled). 

What are the dependent and independent variables?
Dependent Variable: 
Independent Variable: 
```{r}
# perform one-way ANOVA here!
```
What was the p-value?

What does this tell us about the relationship between CO2 uptake and treatment? 

## Statistical Test: two way ANOVA
A two-way analysis of variance (ANOVA) is used to determine how two independent variables, in combination, affect a dependent variable. 

While this is a more advanced statistical test, there is no example, just practice! But don't worry, we will walk you through the code! 

### Practice 
The formula used is:
name<-aov(dependent variable ~ independent variable1 * independent variable2, data= data frame)

Based on what we have been doing, what are our dependent and independent variables?
Dependent: 
Independent 1:
Independent 2:
*note that independent variables 1 and 2 can be interchangeable! 
```{r}
try<-aov(uptake ~ Treatment * Type, data= CO2)
summary(try)
```
What was the p-value for treatment? 

What was the p-value for type? 

What was the p-value for the interaction between treatment and type? 

What is a two-way ANOVA used for? What do these results tell us about our two independent variables? 








